Supplementary material for "Hack Weeks as a model for Data Science Education and Collaboration"

### 2017-12-12 supplementary_material.tex
geohack facilitation

The hack weeks presented here used different approaches, that nevertheless several common themes and ideas. In particular, they focus on making the selection procedure as quantitative as possible and use the data collected during each stage of the process to critically evaluate the efficacy of the process and its subject to human bias.
Secondly, much energy is expended early on in the design of the application form, which often use a combination of open-answer and multiple-choice questions to assess candidates' suitability, skill levels and backgrounds. Care must be taken when designing questions assessing skill levels, because options such as ``beginner'' or ``expert'' are have no absolute scaling and may have different meanings to different applicants. They may also be subject to intrinsic biases, if candidates of certain backgrounds systematically underestimate or overestimate their skill levels compared to other candidates (the Dunning Kruger effect). It is thus advisable to calibrate skill levels on concrete milestones. For example, instead of asking participants whether they are ``beginners'' or ``experts'' at machine learning, we instead asked them whether they had done machine learning for class work, had used techniques from machine learning in their own research, or had developed and implemented their own methods. Similarly, questions about programming skills may ask whether participants can perform a particular task in one of a list of languages  (e.g., "write a module or a library"). Responding to questions about demographic characteristics should generally be optional, and organizers should consider how these questions might interact adversely with workshop goals, e.g.\ via stereotype threat. For example, Astro Hack Week only asked whether an applicant considered themselves a minority with respect to gender identity within astronomy, rather than asking them to explicitly self-identify within a specific category. Meanwhile, neuro hack week opted to ask about gender ethnicity and race, using questions that are standardly used for example, when collecting statistics about study participants.

@@

Some of our hack week tutorials follow the structure of the Software Carpentry model \cite{b:wilson-swc-lessons-2016}. The content is divided in to a series of lessons, each having a well-defined set of learning objectives, key questions, and expected outcomes. Interspersed throughout each lesson are a series of exercises which participants work through at their own pace. We frequently use a method drawn from Software Carpentry whereby students indicate their need for help on an exercise via posting a colored sticky-note on their computer. This provides a simple method to deliver help to those students who need it without interrupting the flow of the tutorial. We encourage instructors to develop online material to accompany their tutorials, either in the form of Jupyter Notebooks \cite{kluyver2016jupyter} or following the markdown templates provided by Software Carpentry. This way the hack week "curriculum" can be built up over time and students can refer to online content after the tutorial if they need to revisit certain topics. Depending on the content, some instructors convey concepts during the tutorial via "live-coding" while others display static code and their results. Either way, we find divding code into short blocks, each with a specific task, helps participants from becoming saturated with informaiton.
In some cases, tutorials require substantial software installation and data. Though we encourage participants to install these software dependencies on their own computers, so that they can continue using them after the tutorial, this is sometime so complicated as to slow down teaching, and hold back. In some cases, we have experimented using \emph{Jupyterhub}\footnote{https://jupyterhub.readthedocs.io/en/latest/} to provide students with immediate access to computational environment that provides teaching materials (e.g., Jupyter notebooks), software dependencies, and data \cite{holdgraf2017portable}. Providing access to an ephemeral consistent environment for the purpose of teaching has become even simpler with the recent advent of the \emph{Binder} system \cite{Holdgraf2017-pd, Titus_Brown_undated-pc}

@@

@@

The \textit{impostor phenomenon}, or \textit{impostor syndrome} (IS) is a dissonant feeling experienced by certain high-achieving individuals, that despite objective evidence to the contrary, they are in fact not as intelligent or capable as they appear.
Individuals with IS thus experience a fear of being ''found out``, shamed and expelled from their environment \cite{Clance1978-ef}.
Initial observations of the first Astro Hack Week conducted by data science ethnographer Brittany Fiore-Gartland\footnote{\url{http://astrohack week.org/blog/ethnographic-notes.html}} suggested that hack weeks are an environment prone to a particular kind of IS: participants might feel the need to be experts in multiple aspects of the activities pursued during a hack week: expertise in a scientific domain, as well as expertise in a the variety of technical tools used.
This particular form of IS hinges to some degree on the design focus on diversity of backgrounds (everyone else seems to know something that you do not!) and might be further exacerbated by the expectation that attendants expose their ideas to public scrutiny, find collaborators in a very short amount of time, and not least produce and present a successful hack at the end of the week\footnote{see also this insightful blog post: \url{https://medium.com/astronomy-without-stars/the-horror-of-hack-days-52c6b52cfc3b}}.
The prevalence of IS at a hack week may be endemic to the format, and should thus be a major concern for any organizing committee.
This is because of the chilling effect it tends to have on participants and the community as a whole, and particularly on women (in particular in fields in which women are under-represented) and members of ethnic and racial minorities, correlating with anxiety and other forms of mental distress \cite{Parkman2016-ro}.
Less severely, another major concern is that IS inhibits risk taking: participants experiencing it will be less likely to ask a question, to put forward an idea for a hack, to be pro-active about forming new collaborations.
 Many of the goals of a hack week, including the successful completion of projects, lateral knowledge transfer, as well as community building are hampered.
We are working within our hack weeks to mitigate IS using various techniques.
With respect to minority participants, ensuring adequate representation can decrease feelings of otherness and may help reduce IS.
More generally, being open about the presence and prevalence of IS can help participants feel more at ease.%(e.g. one Astro Hack Week participant remarked in a survey, \textit{''I really appreciated the direct acknowledgement of impostor syndrome on the first day.
%I think it helped ease the feeling!!``}).
Additionally, role models are very effective at encouraging positive behaviour: asking participants with prior experience at hack weeks at all academic levels to ask questions, even when they might know the answer, can help foster an inclusive environment that rewards risk taking.

>>>>

The hack weeks presented here used different approaches, that nevertheless several common themes and ideas. In particular, they focus on making the selection procedure as quantitative as possible and use the data collected during each stage of the process to critically evaluate the efficacy of the process and its subject to human bias.
Secondly, much energy is expended early on in the design of the application form, which often use a combination of open-answer and multiple-choice questions to assess candidates' suitability, skill levels and backgrounds. Care must be taken when designing questions assessing skill levels, because options such as ``beginner'' or ``expert'' are have no absolute scaling and may have different meanings to different applicants. They may also be subject to intrinsic biases, if candidates of certain backgrounds systematically underestimate or overestimate their skill levels compared to other candidates (the Dunning Kruger effect). It is thus advisable to calibrate skill levels on concrete milestones. For example, instead of asking participants whether they are ``beginners'' or ``experts'' at machine learning, we instead asked them whether they had done machine learning for class work, had used techniques from machine learning in their own research, or had developed and implemented their own methods. Similarly, questions about programming skills may ask whether participants can perform a particular task in one of a list of languages  (e.g., ``write a module or a library''). Responding to questions about demographic characteristics should generally be optional, and organizers should consider how these questions might interact adversely with workshop goals, e.g.\ via stereotype threat. For example, Astro Hack Week only asked whether an applicant considered themselves a minority with respect to gender identity within astronomy, rather than asking them to explicitly self-identify within a specific category. Meanwhile, neuro hack week opted to ask about gender ethnicity and race, using questions that are standardly used for example, when collecting statistics about study participants.

@@

Some of our hack week tutorials follow the structure of the Software Carpentry model \cite{b:wilson-swc-lessons-2016}. The content is divided in to a series of lessons, each having a well-defined set of learning objectives, key questions, and expected outcomes. Interspersed throughout each lesson are a series of exercises which participants work through at their own pace. We frequently use a method drawn from Software Carpentry whereby students indicate their need for help on an exercise via posting a colored sticky-note on their computer. This provides a simple method to deliver help to those students who need it without interrupting the flow of the tutorial. We encourage instructors to develop online material to accompany their tutorials, either in the form of Jupyter Notebooks \cite{kluyver2016jupyter} or following the markdown templates provided by Software Carpentry. This way the hack week ``curriculum'' can be built up over time and students can refer to online content after the tutorial if they need to revisit certain topics. Depending on the content, some instructors convey concepts during the tutorial via ``live-coding'' while others display static code and their results. Either way, we find divding code into short blocks, each with a specific task, helps participants from becoming saturated with informaiton.
In some cases, tutorials require substantial software installation and data. Though we encourage participants to install these software dependencies on their own computers, so that they can continue using them after the tutorial, this is sometime so complicated as to slow down teaching, and hold back. In some cases, we have experimented using \emph{Jupyterhub}\footnote{https://jupyterhub.readthedocs.io/en/latest/} to provide students with immediate access to computational environment that provides teaching materials (e.g., Jupyter notebooks), software dependencies, and data \cite{holdgraf2017portable}. Providing access to an ephemeral consistent environment for the purpose of teaching has become even simpler with the recent advent of the \emph{Binder} system \cite{Holdgraf2017-pd, Titus_Brown_undated-pc}

@@

At geohack week we employed ``Liberating Structures'', a facilitation approach in which we introduce small shifts in the way groups engage in order to increase relational coordination and trust. We invited a professional facilitator to assist us in leading a series of conversations and activities aimed at fostering group engagement. In one activity, participants faced one another and were given a brief statement inviting a two minute conversation, after which the rows rotated to maximize the mixing of ideas. Invitations included statements such as ``I belong at geohack because...'' or ``A challenge I hope we can make progress on is...''. Importantly, event organizers joined in the activity, which we found helped participants feel as if they were not alone in facing some of the challenges inherent in data science studies. Another activity was designed to address the challenges of forming project teams. Individuals were invited to briefly pitch project ideas to the full group, and then form stations around the room. The remaining participants were then invited to visit any station that drew their interest, and to rotate between up to two other stations if they desired. We found this approach minimized the pressure on participants, giving them time to get a feel for project ideas and group dynamics before joining a team.

@@

The \textit{impostor phenomenon}, or \textit{impostor syndrome} (IS) is a dissonant feeling experienced by certain high-achieving individuals, that despite objective evidence to the contrary, they are in fact not as intelligent or capable as they appear.
Individuals with IS thus experience a fear of being ``found out'', shamed and expelled from their environment \cite{Clance1978-ef}.
Initial observations of the first Astro Hack Week conducted by data science ethnographer Brittany Fiore-Gartland\footnote{\url{http://astrohack week.org/blog/ethnographic-notes.html}} suggested that hack weeks are an environment prone to a particular kind of IS: participants might feel the need to be experts in multiple aspects of the activities pursued during a hack week: expertise in a scientific domain, as well as expertise in a the variety of technical tools used.
This particular form of IS hinges to some degree on the design focus on diversity of backgrounds (everyone else seems to know something that you do not!) and might be further exacerbated by the expectation that attendants expose their ideas to public scrutiny, find collaborators in a very short amount of time, and not least produce and present a successful hack at the end of the week\footnote{see also this insightful blog post: \url{https://medium.com/astronomy-without-stars/the-horror-of-hack-days-52c6b52cfc3b}}.
The prevalence of IS at a hack week may be endemic to the format, and should thus be a major concern for any organizing committee.
This is because of the chilling effect it tends to have on participants and the community as a whole, and particularly on women (in particular in fields in which women are under-represented) and members of ethnic and racial minorities, correlating with anxiety and other forms of mental distress \cite{Parkman2016-ro}.
Less severely, another major concern is that IS inhibits risk taking: participants experiencing it will be less likely to ask a question, to put forward an idea for a hack, to be pro-active about forming new collaborations.
 Many of the goals of a hack week, including the successful completion of projects, lateral knowledge transfer, as well as community building are hampered.
We are working within our hack weeks to mitigate IS using various techniques.
With respect to minority participants, ensuring adequate representation can decrease feelings of otherness and may help reduce IS.
More generally, being open about the presence and prevalence of IS can help participants feel more at ease.%(e.g. one Astro Hack Week participant remarked in a survey, \textit{''I really appreciated the direct acknowledgement of impostor syndrome on the first day.
%I think it helped ease the feeling!!``}).
Additionally, role models are very effective at encouraging positive behaviour: asking participants with prior experience at hack weeks at all academic levels to ask questions, even when they might know the answer, can help foster an inclusive environment that rewards risk taking. We also invited our organizing committees to model open communication about their own handling of IS through the course of their career. We found that transparency about one's own uncertainty, espcially from community leaders, helped create a hospitable atmosphere and break down barriers.
